Boosting Higgs Pair Production in the $bb\bar{b}\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-07-09 paper/sec-analysis.tex
work on paper

>>>>

\subsection{Overlap between different categories}
\label{sec:overlap}

As has been discussed in Sect.~\ref{sec:analysis}, while the optimization of each
category is performed in an inclusive analysis, after having determine the order
of significance of the categories we turn to a exclusive analysis, in order that
we can combine consistently the results from the various topologies.
%
In this appendix we discuss the role of the overlap
between  different categories, which is of course ignored in the final exclusive
analyses.
%
The idea is to understand how significant the information is on which events satisfy the requirements
of more than one category at the same time.

To illustrate the role of the overlap between the different categories, in
Fig.~\ref{fig:categorisationHisto} we show the fraction of events that satisfy the requirements
of one or more categories, for signal and background events.
%
We consider the following cases
\begin{itemize}
\item Events that satisfy the requirements of the resolved category only,
\item boosted category only
\item Intermediate category only
\item Both the resolved and the intermediate category
  \item Both the boosted and the intermediate category
\end{itemize}
As we can see from Fig.~\ref{fig:categorisationHisto}, an overwhelming majority of the events
below to the resolved category, specially in the background case.
%
Signal events are dominant in both the intermediate and the resolved categories.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.65\textwidth]{plots/overlap_categories_C1.pdf}
\caption{\small The fraction of events that satisfy the requirements
  of one or more categories, for signal and background events.
}
\label{fig:categorisationHisto}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

bin 0: resolved only
bin 1: intermediate only
bin 2: boosted only
bin 3: resolved-intermediate overlap
bin 4: resolved-boosted overlap
bin 5: intermediate-boosted overlap (should be empty due to orthogonal jet multiplicity cuts)

### 2015-07-09 paper/sec-intro.tex
work on paper

The measurement of double Higgs production will be one of the main
physics goals of the LHC program in its recently started high-energy
phase, as well as for its future high-luminosity phase, as
well as of the program of any future high-energy collider.
%
Double Higgs production is directly sensitive to the Higgs trilinear coupling, and thus 
provides information on the scalar potential responsible for electroweak symmetry breaking.
%
It is also sensitive to the underlying strength of the Higgs interactions
at high energies, and can test the composite nature of the 
Higgs boson~\cite{Giudice:2007fh,Contino:2010mh}.

In the Standard Model (SM), the dominant mechanism for the production of two Higgs bosons at the LHC is 
gluon fusion (see Ref.~\cite{baglio} and references therein), analogously to single Higgs production.
For a center-of-mass energy $\sqrt{s} = 14\,$TeV, the recently computed next-to-next to
leading order (NNLO)
total cross section is approximately $40\,$fb~\cite{deFlorian:2013jea}.
%
Feasibility studies in the case of a light Higgs boson have been performed for several different final states, including
$b\bar b\gamma\gamma$~\cite{Baur:2003gp,Barger:2013jfa},
$b\bar{b}\tau^+\tau^-$~\cite{Baur:2003gpa,Barr:2013tda,Dolan:2012rv,Dolan:2013rja},
$b\bar{b}W^+W^-$~\cite{Dolan:2012rv,Papaefstathiou:2012qe} and
$b\bar{b}b\bar{b}$~\cite{Baur:2003gpa,Dolan:2012rv,Gouzevitch:2013qca,Cooper:2013kia,Wardrope:2014kya,deLima:2014dta}.
%
While these various studies differ in their qualitative conclusions, it is clear
that the ultimate precision in the determination of the Higgs trilinear
will be provided by the combination of various final states.
%

The advantage of the $b\bar{b}b\bar{b}$ final state is the the signal yield
is enhanced by the large branching fraction of Higgs bosons into bottom-anti-bottom
pairs, ${\rm BR}\lp H\to b\bar{b}\rp\simeq 0.57$~\cite{Dittmaier:2012vm}.
%
However, this channel needs to deal with an overwhelming QCD multi-jet background,
which makes this measurement very challenging.
%
Previous studies of Higgs pair production in this
channel~\cite{Wardrope:2014kya,deLima:2014dta}, considering
only the $4b$ irreducible QCD background, estimate that with a luminosity of
3000 fb$^{-1}$ a signal significance of $S/\sqrt{B}=$AAA can be obtained.
%
However, as show in this work, neglecting other QCD backgrounds such as $bbjj$ and
$jjjj$, with some of these light jets mistagged as $b$-jets, cannot be neglected, making
this measurement even more difficult.

Despite these large backgrounds, by combining the contribution of different final state
topologies, resolved, boosted and intermediate, with a careful optimization
of the analysis strategy, we find that it is possible to obtain a reasonable
good signal discrimination.
%
To further enhance the significance of the measurements, signal and background events
are processed using a Multivariate Analysis (in particular a feed-forward neural
network), that further enhance $S\sqrt{B}$ up to a value of AAA.
%
Therefore, we are able to rescue the potential of this final state in a realistic
analysis setup.

One of the original motivations of this study was to provide guidance to the experiments
concerning which aspects of the measurement should be particularly optimized
to increase the signal significance in this channel.
%
In this respect, the study the dependence on our results on the $b$-tagging mistag
efficiency and the light and charm jet mistag rate, the jet energy and transverse
momentum resolution, or the Higgs invariant mass resolution.
%
In each of these cases, we determine the significance  $S/sqrt{B}$ of the
measurement in both conservative and optimistic assumptions of these
parameters.

The structure of this paper is the following.
%
In Sect.~\ref{mcgeneration} we discuss the generation of the signal
and background Monte Carlo samples.
%
Then in Sect.~\ref{sec:analysis}
we describe our analysis strategy, and in particular
the classification of each individual event into
different categories according to the event topology.
%
The results for our traditional cut-based analysis
are presented in Sect.~\ref{sec:results}, and in
Sect.~\ref{sec:mva} we show how we can enhance signal
significance by means of multivariate techniques.
%
Finally in Sect.~\ref{sec:conclusions} we conclude and outline
possible future developements for Higgs pair production
in the $b\bar{b}b\bar{b}$ final state.

>>>>

The measurement of double Higgs production will be one of the main
physics goals of the LHC program in its recently started high-energy
phase, as well as for its future high-luminosity upgrade (HL-LHC),
which aims to accumulate a total integrated
luminosity of 3 ab$^{-1}$~\cite{ATLAS:2013hta}.
%
Higgs pair production is directly sensitive to the
Higgs trilinear coupling $\lambda$~\cite{baglio}, and thus 
provides crucial
information on the electroweak symmetry breaking mechanism.
%
It is also sensitive to the underlying strength of the Higgs interactions
at high energies, and can be used to test the composite nature of the 
Higgs boson~\cite{Giudice:2007fh,Contino:2010mh}.
%
In addition, Higgs pair production will also be a central process at
any future high-energy 
collider, either at a 100 TeV circular machine~\cite{Barr:2014sga} or at
a linear or circular electron-positron collider~\cite{Contino:2013gna}.

In the Standard Model (SM), the dominant mechanism for the production of two Higgs bosons at the LHC is 
gluon fusion (see Ref.~\cite{baglio} and references therein), analogously to single Higgs production~\cite{Dittmaier:2012vm}.
%
For a center-of-mass energy $\sqrt{s} = 14\,$TeV, the next-to-next-to-leading order (NNLO)
total cross section is approximately $40\,$fb~\cite{deFlorian:2013jea},
which is increased by a further few percent once NNLL corrections
are accounted for~\cite{deFlorian:2015moa}.
%
Feasibility studies in the case of a SM-like Higgs boson
in gluon fusion have been performed for several different final states, including
$b\bar b\gamma\gamma$~\cite{Baur:2003gp,Barger:2013jfa},
$b\bar{b}\tau^+\tau^-$~\cite{Baur:2003gpa,Barr:2013tda,Dolan:2012rv,Dolan:2013rja},
$b\bar{b}W^+W^-$~\cite{Dolan:2012rv,Papaefstathiou:2012qe} and
$b\bar{b}b\bar{b}$~\cite{Baur:2003gpa,Dolan:2012rv,Wardrope:2014kya,deLima:2014dta}.
%
While these studies differ in their qualitative conclusions,
a consistent picture emerges 
that the ultimate precision in the determination of the Higgs trilinear
will only be achieved from the combination of various final states.

In addition to the gluon-fusion production mode, Higgs pairs
can be also produced in the vector-boson fusion
channel $hhjj$~\cite{Contino:2010mh,Dolan:2013rja,Dolan:2015zja},
the associated production modes $hhW$ and $hhZ$, and also in association
with top quark pairs $hht\bar{t}$~\cite{Englert:2014uqa}.
%
All these channels are extremely challenging due to the small production
rates: at 14 TeV, the inclusive total cross-sections are
2 fb for VBF $hhjj$~\cite{Liu-Sheng:2014gxa},
0.5 fb for $W(Z)hh$~\cite{baglio}
and 1.0 for $hht\bar{t}$~\cite{Englert:2014uqa}.
%
Concentrating on the SM production dynamics of Higgs pair
production in the various channels, several studies
have investigate the interplay of various kinematical
distributions towards the optimisation of the
extraction of $\lambda$ from the measured
cross-sections, as well as the improvement
in the signal modeling taking into account
higher-order corrections and theoretical
uncertainties~\cite{Slawinska:2014vpa,Chen:2014xra,Goertz:2013kp,Frederix:2014hta,Dawson:2015oha,Maltoni:2014eza}
%

While the SM production rates for Higgs
pair production are small in all production modes,
and require the full integrated luminosity of the HL-LHC phase
to be accessed, double Higgs production rates can be substantially
enhanced in a variety of Beyond the SM (BSM) scenarios.
%
Feasibility studies of Higgs pair production in New Physics
models have been performed in a number of different frameworks,
including effective theories with higher-dimensional
operators and anomalous values of the
Higgs couplings~\cite{Nishiwaki:2013cma,Dall'Osso:2015aia,Azatov:2015oxa,Liu:2014rba,Goertz:2014qta,He:2015spf,Grober:2015cwa}, resonant production
in models such as extra dimensions~\cite{Gouzevitch:2013qca,Cooper:2013kia,No:2013wsa},as well as
in the context of Supersymmetry or Two Higgs Doublet models (2HDMs)~\cite{Belyaev:1999kk,Han:2013sga,Hespel:2014sla,Wu:2015nba,Cao:2014kya,Ellwanger:2013ova}.

%
In many of these scenarios, the production rates are enhanced
and Higgs pair production becomes accessible already at
Run II, while still being consistent with current limits on BSM searches
and single Higgs coupling measurements.
%
With this motivation, searches for the production of Higgs pairs, whose rates
could be enhanced by BSM effects have already been performed
by ATLAS in the $b\bar{b}b\bar{b}$~\cite{Aad:2015uka}
and $b\bar{b}\gamma\gamma$~\cite{Aad:2014yja} final states,
and by
and CMS in the same $b\bar{b}b\bar{b}$~\cite{Khachatryan:2015yea}
and $b\bar{b}\gamma\gamma$~\cite{Chatrchyan:2011wt} final
states.
%
though the sensitivity of these searches is still far from
that required for a SM production mechanism.
%
However, one should note that in general BSM dynamics modify substantially
the kinematic distributions of the Higgs decay products, for
instance boosting the di-Higgs system, and therefore
different analysis strategies would be requires in BSM
Higgs pair searches as compared to SM production measurements.

In the context of SM production,
the main advantage of the $b\bar{b}b\bar{b}$ final state is the the signal yield
is enhanced by the large branching fraction of Higgs bosons into bottom-anti-bottom
pairs, ${\rm BR}\lp H\to b\bar{b}\rp\simeq 0.57$~\cite{Dittmaier:2012vm}.
%
However, a measurement in this channel
needs to deal with an overwhelming QCD multi-jet background,
which makes it very challenging.
%
Previous studies of Higgs pair production in this
channel~\cite{Wardrope:2014kya,deLima:2014dta}
estimate that with a luminosity of
3000 fb$^{-1}$ a signal significance of around $S/\sqrt{B}\sim 2.0$ can be obtained.
%
Both studies were based on exploiting the differences between signal and background
using techniques that account for a large number of kinematic
variables, Boosted Decision Trees
and Shower Reconstruction~\cite{Soper:2011cr} respectively.
%
Concerning the QCD multijet backgrounds, Refs.~\cite{Wardrope:2014kya,deLima:2014dta} the light-jet processes $bbjj$ and
$jjjj$,
which can fake the signal when 
some of the light jets are mistagged as $b$-jets.
%
As will show, this is not a good approximation since both due to
combinatorics and to $b$-quark radiation in the shower, the
contribution of $2b2j$ and $4j$ can become comparable
to that of $4b$.

In this work we revisit the feasibility of SM Higgs pair production by gluon-fusion
in the $4b$ final state.
%
Our analysis is based on exploiting the differences between a large
number of kinematic variables in signal and background events.
%
Despite these large QCD multijet backgrounds, by combining the contribution of different final state
topologies, resolved, boosted and intermediate, with a careful optimization
of the analysis strategy, we find that it is possible to obtain a reasonable
good signal discrimination.
%
To further enhance the significance of the measurements, signal and background events
are processed using a Multivariate Analysis (in particular a feed-forward neural
network), that further enhance $S/\sqrt{B}$ up to a value of slightly above 3.0
%
Therefore, we are able to rescue the potential of this final state in a realistic
analysis setup.

One of the original motivations of this study was to provide guidance to the experiments
concerning which aspects of the measurement should be particularly optimized
to increase the signal significance in this channel.
%
In this respect, the study the dependence on our results on the $b$-tagging mistag
efficiency and the light and charm jet mistag rate, the jet energy and transverse
momentum resolution, or the Higgs invariant mass resolution.
%
In each of these cases, we determine the significance
$S/\sqrt{B}$ of the
measurement in both conservative and optimistic assumptions of these
parameters.

The structure of this paper is the following.
%
In Sect.~\ref{mcgeneration} we discuss the generation of the signal
and background Monte Carlo samples.
%
Then in Sect.~\ref{sec:analysis}
we describe our analysis strategy, and in particular
the classification of each individual event into
different categories according to the event topology.
%
The results for our traditional cut-based analysis
are presented in Sect.~\ref{sec:results}, and in
Sect.~\ref{sec:mva} we show how we can enhance signal
significance by means of multivariate techniques.
%
Finally in Sect.~\ref{sec:conclusions} we conclude and outline
possible future developments for Higgs pair production
in the $b\bar{b}b\bar{b}$ final state.
